Are Delayed Issues Harder to Resolve?

### 2016-06-17 phaseDelay4Emse.tex
Updates from ShareLaTeX

\section{Introduction}
In 2013-2014, 
eleven  million programmers~\cite{avram14} and
half a trillion dollars~\cite{pettey14} were spent on information technology.
Such a large and growing effort should be managed and optimized via  well-researched conclusions.

It is standard practice
in other fields, such as medicine,
to continually revisit old conclusions~\cite{prasad13}.
Accordingly, this paper revisits
the commonly held belief we label {\em delayed issue effect} (hereafter, DIE):
% Later in this paper, we offer a precise definition for this effect.
% For the moment, we describe it as follows:
more effort is required to resolve an issue the longer an issue lingers in a system.
% it is far easier to resolve issues earlier rather than later in the lifecycle.
 \fig{b81} shows an example of the delayed issue effect (relating
 the relative cost of fixing requirements issues at different phases of a project).

 Basili \& Boehm comment that, since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!b]  
\begin{center}
\includegraphics[angle=270,width=4in]{img/b81.pdf}
 \end{center}
 \caption{An example DIE effect. From Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}
Other prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.  McConnell mentions it as a "common observation" in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this effect, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that that the delayed issue effect is a generally useful law of software engineering.
For example, what we call the delayed issued effect was listed first by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

In analyzing data from a contemporary set of software development projects, however, we did not find results to corroborate these claims. While the delayed issue effect might have been a dominant
effect decades ago, perhaps not so for $21^{\mathit{st}}$
software development. 
 The delayed issue effect was first reported in 1976 in a era of punch card programming
and non-interactive environments~\cite{Boehm76}. In the 21$^\mathit{st}$ century, we  program in 
interactive environments with higher-level languages and better source code control
tools. Such tools allow for the faster refactoring of existing
code-- in which case, 
managing the changes required to fix (say) an incorrect requirements assumption
is far less    onerous    than before. Further, software engineering theory and practice has evolved into new paradigms focused on rapid feedback and delivery, enabled by significant technological advances in the past 40 years. There is little empirical evidence for the delayed issue effect since its initial observation, no doubt due in part to DIE being ``just common sense'' as Glass states~\cite{glass02}.

This article explores the currency of the delayed issue effect.
% Given all of the changes to development practice in the time since the effect was first observed, finding no softening of the DIE might be taken as more surprising than our current results. These changes in software development practice are discussed later in this paper.
% The above argument is an anecdotal evidence that  the delayed issue effect might no longer exist. But anecdotes are not rigorous
% evidence. Accordingly,  this article explores the currency of the delayed issue effect.
After some initial definitions, we discuss the
value of checking old ideas. Next, we present a survey of industrial practitioners and researchers that documents the widespread belief that delayed issues have a negative impact on projects.  After that, we  analyze 171 software  projects developed in the period 2006--2014 and find  {\em no evidence} of the delayed issue effect. Finally, we discuss the validity and implications of our results, as well as possible reasons for the lack of observed effect given the state of the practice. To ensure reproducibility,
all the data  used in this study is available in the PROMISE
repository at openscience.us/repo. 
To the best of our knowledge,
this the largest study devoted the delayed issue effect yet conducted.

@@

\section{Definitions \& Claims}
\todo{Revisit these definitions after reviewing the rest of the paper for consistency of terms}
This paper uses the following definitions:
\bi
\item
The {\em delayed issue effect}:   it is {\em very much}  more {\em difficult} to resolve  issues in a software project, the {\em longer} they remain.
\item
 {\em Longer} time is defined as per  Boehm'81~\cite{Boehm81}; i.e. the gap between the   phases where   issues are introduced and resolved.
\item
We say that a measure $m$ collected in phase ${1,.,i,..j}$ is 
{\em very much} more when  that
   measure at phase $j$   
   is larger than the sum of those measures in the earlier phases;
   i.e. $\sum_{i=1}^{j-1} m_i $. 
\item
Issues are more {\em difficult}  
when their resolution takes more time or costs more  (e.g. needs expensive
debugging tools or the skills of expensive developers).
\ei
% Note that we use  the  term ``delayed issue effect'' as generalization of the
% more specific rule  ``requirements errors are hardest to fix''.
% This generalization is valid since the rationale for the rule about requirements
% is usually done as per McConnell~\cite{mcconnell01}; i.e. small upstream mistakes very
% early in the system can cascade into huge problems later in the lifecycle.
% That said, we prefer our more general term ``delayed issue effect'' since not
% only might requirements errors cascade, so too might analysis errors, design errors, etc.

@@

This paper defends the  following two claims and one hypothesis. Note hypotheses are defended
using some statistical
significance test while the claims are  supported via a variety
of arguments presented later in this paper.

{\bf  Claim1: ``DIE'' is a  commonly held belief.}
Using a literature review, we confirm that there are numerous historical papers (dating back decades)
that endorse the DIE. Also, using a survey of practitioners and researchers, we find  that 
 the DIE  appears as the
most strongly-held belief amongst software engineers.

{\bf  Claim2: ``DIE'' is poorly documented.}
 As discussed in our literature review,  many of the papers reporting the DIE
 either (1)~are quite old (papers dating from last century);
(2)~quote prior papers without presenting   new data; 
(3)~or cite data sources that can no longer be
confirmed.

@@

\section{ Reassessing Old Truisms}
Beliefs in general
principles of software
engineering are common to both research and practice. Professional societies assume such generalities exist when they offer
 lists of supposedly general ``best practices'' such as
the IEEE 1012 standard for software verification~\cite{1012}. 
 Endres \& Rombach offer empirical observations, theories, and laws\footnote{Endres \& Rombach note that these are not laws of nature in the scientific sense, but theories with repeated empirical evidence.}~\cite{endres03}.
 Many other 
commonly cited researchers  do the same, e.g.,
Glass~\cite{glass02}, Jones \cite{jones07}, and Boehm~\cite{boehm00b}.
Budgen \& Kitchenham seek to reorganize SE research using
general
conclusions drawn from a larger number of studies~\cite{kitch04,budgen09}.

@@

The cognitive issues that complicate medicine are also found in software engineering.
Passos et al.~\cite{passos11} warn that developers
usually assume that the truisms they learn from a few past
projects are general to 
all their future projects. They comment ``past experiences were taken into account without 
much consideration for their context''~\cite{passos11}.
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} support Passos et al. In an empirical study of expert effort estimation, they report that the experts rarely use lessons
  from past projects to improve their future reasoning in effort estimation~\cite{jorgensen09}. 
 They note that,
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).
Just as in medicine, 
%our field suffers when
 %software engineers do  not revise old beliefs.  Therefore, 
 it is important for our field
 to regularly  reexamine    old beliefs such as the delayed issue effect.

\section{``DIE'' is a  commonly held belief}
\label{sect:belief}
% To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
% understanding of SE theory, then the DIE  may well influence their decisions.

We conducted two surveys of software engineers to assess the prevalence of DIE. 
The surveys collected data on software engineers' views of DIE and other commonly held software engineering ``laws''. The surveys were conducted using Amazon's Mechanical Turk. The first survey was conducted only with professional software engineers. Participants were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology. The  second survey was conducted with Program Committee members of the ESEC/FSE 2015 and ICSE 2014 conferences solicited via email.

The practitioner survey presented the following law: ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}). We abbreviate this law as RqtsErr.\footnote{We use the RqtsErr formulation since this issue typically needs no supportive explanatory
text. If we had asked respondents about our more general term ``delayed issue
effect'', we would have had to burden our respondents with extra explanations.}
The PC member survey presented the RqtsErr law and an additional law on the DelayedIssueEffect: ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix''.

% Laws other than RqtsErr are not relevant to this paper, but we add them to as comparison points.

The respondents answered two questions in response to each law:
\bi
\item
\textbf{Agreement:} ``Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree. A text box was provided to explain the answer. \newline
\item
\textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were:
\bi
\item -1: I don't know
\item 0:  this law does not apply at all
 \item 1: applies in a very narrow range of projects  
 \item 2: rarely applies
 \item 3: occasionally applies
\item 4: very frequently applies
\item 5: always applies
\ei
Respondents were required to explain the applicability score in a text box.
\ei

\begin{figure}[!ht] 
\scriptsize 
 \begin{center}
\begin{tabular}{l|c|c|c|c|c}
 &  & \multicolumn{2}{c|}{agreement} & \multicolumn{2}{c}{applicability} \\
Practitioner survey  & N & med & mode & med & mode \\
\hline 
\textbf{Rqts errors are most expensive...} & 16 & 5 & 5 & 4 & 5 \\ 
Inspections can remove 90\% of defects & 18 & 4 & 5 & 4 & 5 \\
80-20 rule (defects to modules) & 12 & 4 & 5 & 4 & 5 \\
Most time is spent removing errors & 16 & 4 & 4 & 4 & 5 \\ 
Process maturity improves output & 17 & 4 & 4 & 4 & 4 \\ 
Missing reqts are hardest to fix & 17 & 4 & 4 & 4 & 4 \\
Reuse increases prod. and qual. & 16 & 4 & 4 & 4 & 4 \\
OO-programming reduces errors & 13 & 4 & 4 & 4 & 3 \\
Adding manpower to a late project & 15 & 4 & 4 & 4 & 4 \\
Smaller changes have higher error density & 14 & 3 & 3 & 3.5 & 5 \\
A developer is unsuited to test own code & 17 & 3 & 1 & 4 & 5\\
 \multicolumn{6}{l}{}\\
\multicolumn{6}{l}{Researcher survey} \\\hline 
Process maturity improves output & 4 & 4 & 4 & 4 & 5 \\
\textbf{Rqts errors are most expensive...} & 30 & 4 & 4 & 4 & 4   \\ 
\textbf{DelayedIssueEffect} & 30 & 4 & 4 & -- & --  \\
Reuse increases prod. and qual. & 6 & 4 & 4 & 4 & 4 \\
80-20 rule (defects to modules) & 6 & 4 & 4 & 4 & 3 \\
Missing reqts are hardest to fix & 7 & 4 & 4 & 4 & 3 \\
OO-programming reduces errors & 6 & 4 & 4 & 3 & 4 \\
Inspections can remove 90\% of defects & 7 & 4 & 4 & 3 & 3 \\
Adding manpower to a late project & 4 & 3 & 4 & 4 & 3 \\
Most time is spent removing errors & 6 & 3 & 3 & 4 & 4 \\ 
Smaller changes have higher error density & 4 & 3 & -- & 4 & 4 \\
A developer is unsuited to test own code & 7 & 2 & 1 & 3 & 3
\end{tabular} 
 \end{center}
\caption{Agreement and applicability of SE axioms.}
\label{fig:survey_results}
\end{figure}

Both  practitioners and researchers strongly believed in RqtsErr. In both sets of responses, RqtsErr received  scores higher than most
other laws. Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers. 
This results strongly support that the notion that the  delayed issue effect   is a commonly held belief in software engineering. From the free response texts, we note that the researchers who disagreed with RqtsErr generally asserted that requirements change can be expensive, but that the effect depends on the process used (e.g., agile vs. waterfall) and the adaptability of the system architecture.

\section{DIE is Poorly  Documented}\label{sect:doco}

One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
in the SE literature. 
%For example,
%we know that \fig{b81} has been presented to the graduate SE class at North Carolina State University, without quarrel or %critical comment, every year for the last decade.
Yet when we look at the literature, the evidence for
delayed issue effect is both very sparse and very old.
The first data on the difficulties of resolving delayed issues as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}). These studies are most often cited by secondary sources on the delayed issue effect (e.g., \cite{boehm01, mcconnell01,glass02}. We note that it is unclear from the text in ~\cite{Daly77} and \cite{Boehm76} if cost is defined in terms of effort, or in actual cost (i.e., labor, materiel, travel, etc). The data points from these studies are not published for analysis. These seminal studies are the basis for most DIE claims and that the cost to find and fix a defect monotonically increases with lifecycle phase.

\begin{figure}[!b]
\begin{center}\includegraphics[width=4in]{img/boehm_cost-to-fix.png}\end{center}
 \caption{Historical cost-to-fix curve. From~\cite{Boehm81}.}\label{fig:cost-to-fix}
 \end{figure}

In the 40 years since these initial studies, few studies have explored the difficulty to resolve issues
as a function of lifecycle phase. 
Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects. Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but the raw data points were not provided and thus cannot be confirmed.

\todo{REview the primary sources for this paper}
This was a common theme in the literature reviewed for this paper-- i.e.  that  it was no longer possible to access
the data used to make prior conclusions.
As an example of this, \fig{steck} shows one 2004 survey that reports an 
greatly increased delayed issue  effect for 
requirements issues in eight case studies. Note that we cannot verify
any of those results since the links in the references of
that  survey are all broken (``page not found'' errors).

A goal of agile methods 
is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}. Relatively little empirical data exists on this point.
Elssamadisy and Schalliol~\cite{Elssamadisy02} offer an anecdotal report on the growing, high cost of rework in a 50 person, three-year, 500KLOC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their 
exact figures.

@@

 \begin{figure}
{\small
\begin{center}
\begin{tabular}{r|rrrr}
 case& \multicolumn{4}{c}{Phase Requirements Issue Found }  \\
 study               &Requirements & Design & Code&  Test\\\hline
1& 1 &3& 5& 37\\  
2& 1  &    & 10 & 40 \\ 
3& 1   &    & 10 &  40 \\
4&  1  &   5 &      & 50  \\
5&  1 &3& 7& 51 \\
6& 1& 5 &33 &75  \\
7&  1 & 20 & 45 & 250 \\  
\end{tabular}
\end{center}}
\caption{Cost to resolve requirements issues, relative to resolving  during requirements. From ~\cite{steck04}.}\label{fig:steck}
\end{figure}

@@

Some studies that report less-than large
 increase in the effort required to fix delayed issues. Royce~\cite{Royce98} studied  a million-line, safety-critical missile defense system
(see  Figure~\ref{fig:royce}). Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly. Boehm~\cite{Boehm10} attributes this success to the  development process, which focused on removing architecture risk early in the development lifecycle. Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects ,from two student projects of 2000 deliverable source instructions, is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}). Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to fix certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project). Again, the data points from these studies are not available, and thus the evidence both in favor of and contrary to the DIE is either quite old and/or is based on data that cannot be confirmed.

\begin{figure}
\begin{center}\includegraphics[width=4in]{img/Royce98.png}\end{center}
 \caption{Exception to the rule - The Royce study: cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
 \end{figure}

>>>>

\section{Introduction}
In 2013-2014, 
eleven  million programmers~\cite{avram14} and
half a trillion dollars~\cite{pettey14} were spent on information technology.
Such a large and growing effort should be managed and optimized via  well-researched conclusions.  
To assist in achieving this, there has been a growing recognition within the software engineering research community 
of the importance of theory building ~\cite{Sjoberg08, Paivarinta15, Stol15}. A good theory allows empirical research to go beyond simply
reporting observations and instead provides explanations for why results are observed ~\cite{Stol15}. This occurs by testing theories against data from multiple sources; by reconciling similarities and differences in results it can be determined what factors need to be accounted for in a theory ~\cite{Shull08}. Theory-building needs to be an iterative process,
in which results from practice are used to refine theories and theories are used to inform future observation and data collection ~\cite{Paivarinta15, Stol15}. It is no coincidence that it is standard practice
in other fields, such as medicine,
to continually revisit old conclusions in the light of new theories~\cite{prasad13}.

Accordingly, this paper revisits
the commonly held belief we label the {\em delayed issue effect} (hereafter, DIE):
% Later in this paper, we offer a precise definition for this effect.
% For the moment, we describe it as follows:
more effort is required to resolve an issue the longer an issue lingers in a system.
% it is far easier to resolve issues earlier rather than later in the lifecycle.
 \fig{b81} shows an example of the delayed issue effect (relating
 the relative cost of fixing requirements issues at different phases of a project).

This theory has been used as the basis for decision-making in software engineering. For example,  Basili and Boehm comment that, since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!b]  
\begin{center}
\includegraphics[,width=4in]{img/b81.png}
 \end{center}
 \caption{A widely-recreated chart of the DIE effect. Adapted from Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}

Like any good theory, DIE includes a rationale for why the expected results would be seen. McConnell mentions it as a ``common observation'' in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this rationale, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that that the delayed issue effect is a generally useful law of software engineering.
For example, what we call the delayed issued effect was listed at \#1 by Boehm and Basili in their ``Top 10 list'' of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

In analyzing data from a contemporary set of software development projects, however, we did not find results to corroborate these claims. While the delayed issue effect might have been a dominant
effect decades ago, this does not mean that it is necessarily so for $21^{\mathit{st}}$ century
software development. 
 The delayed issue effect was first reported in 1976 in a era of punch card programming
and non-interactive environments~\cite{Boehm76}. In the 21$^\mathit{st}$ century, we  program in 
interactive environments with higher-level languages and better source code control
tools. Such tools allow for the faster refactoring of existing
code-- in which case, 
managing the changes required to fix (say) an incorrect requirements assumption
is far less    onerous    than before. Further, software engineering theory and practice has evolved into new paradigms focused on rapid feedback and delivery, enabled by significant technological advances in the past 40 years. There is little empirical evidence for the delayed issue effect since its initial observation, no doubt due in part to DIE being ``just common sense'' as Glass states~\cite{glass02}.

This article explores the currency of the delayed issue effect.
% Given all of the changes to development practice in the time since the effect was first observed, finding no softening of the DIE might be taken as more surprising than our current results. These changes in software development practice are discussed later in this paper.
% The above argument is an anecdotal evidence that  the delayed issue effect might no longer exist. But anecdotes are not rigorous
% evidence. Accordingly,  this article explores the currency of the delayed issue effect.
After some initial definitions, we discuss the
value of checking old ideas. Next, we present a survey of industrial practitioners and researchers that documents the widespread belief that delayed issues have a negative impact on projects.  After that, we  analyze 171 software  projects developed in the period 2006--2014 and find  {\em no evidence} of the delayed issue effect. Finally, we discuss the validity and implications of our results, as well as possible reasons for the lack of observed effect given the state of the practice - reasons which, when subjected to further testing, may prove useful for refining the theory. To ensure reproducibility,
all the data  used in this study is available in the PROMISE
repository at openscience.us/repo. 
To the best of our knowledge,
this the largest study devoted the delayed issue effect yet conducted.

@@

\section{Definitions \& Claims}
\label{sect:claims}

This paper uses the following definitions:
\bi
\item
The {\em delayed issue effect}:   it is {\em very much}  more {\em difficult} to resolve  issues in a software project, the {\em longer} they remain.
\item
 {\em Longer} time is defined as per  Boehm'81~\cite{Boehm81}; i.e. the gap between the   phases where   issues are introduced and resolved.
\item
We say that a measure $m$ collected in phase ${1,.,i,..j}$ is 
{\em very much} more when  that
   measure at phase $j$   
   is larger than the sum of those measures in the earlier phases;
   i.e. $\sum_{i=1}^{j-1} m_i $. 
\item
Issues are more {\em difficult}  
when their resolution takes more time or costs more  (e.g. needs expensive
debugging tools or the skills of expensive developers).
\ei
% Note that we use  the  term ``delayed issue effect'' as generalization of the
% more specific rule  ``requirements errors are hardest to fix''.
% This generalization is valid since the rationale for the rule about requirements
% is usually done as per McConnell~\cite{mcconnell01}; i.e. small upstream mistakes very
% early in the system can cascade into huge problems later in the lifecycle.
% That said, we prefer our more general term ``delayed issue effect'' since not
% only might requirements errors cascade, so too might analysis errors, design errors, etc.

@@

This paper defends the following claim and hypothesis. The hypothesis is defended
using some statistical
significance tests while the claim is supported via a variety
of arguments.

{\bf  Claim: ``DIE'' is a  commonly held, yet poorly documented belief.}
We examine the literature promoting the DIE and find that most reference a few primary sources. Many of the papers reporting the DIE
 either (1)~are quite old (papers dating from last century);
(2)~quote prior papers without presenting   new data; 
(3)~or cite data sources that can no longer be
confirmed. We follow-up with a short survey that finds that DIE appears as the most strongly-held belief among software engineers in our sample.

@@

\section{ Reassessing Old Truisms}
General
theories of software
engineering principles are common to both research and practice, although not always explicitly stated. Such theories underlie  lists of proposed general ``best practices'' for effective software development, such as
the IEEE 1012 standard for software verification~\cite{1012}. 
 Endres \& Rombach offer empirical observations, theories, and laws\footnote{Endres \& Rombach note that these are not laws of nature in the scientific sense, but theories with repeated empirical evidence.}~\cite{endres03}.
 Many other 
commonly cited researchers  do the same, e.g.,
Glass~\cite{glass02}, Jones \cite{jones07}, and Boehm~\cite{boehm00b}.
Budgen \& Kitchenham seek to reorganize SE research using
general
conclusions drawn from a larger number of studies~\cite{kitch04,budgen09}.

@@

The cognitive issues that complicate medicine are also found in software engineering.
Passos et al.~\cite{passos11} warn that developers
usually develop their own theories of what works and what doesn't work in creating software, based on experiences from a few past
projects. Too often, these theories are assumed to be general truisms with widespread applicability to future projects. They comment ``past experiences were taken into account without 
much consideration for their context''~\cite{passos11}.
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} support Passos et al. In an empirical study of expert effort estimation, they report that the experts rarely use lessons
  from past projects to improve their future reasoning in effort estimation~\cite{jorgensen09}. 
 They note that,
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).
 A similar effect is reported by
Devanbu et al.~\cite{prem16}  who examined responses from 564 Microsoft software developers from around
the world, they found that  ``(a)~programmers do indeed have very
strong beliefs on certain topics; (b)~their beliefs are primarily formed
based on personal experience, rather than on findings in empirical
research; (c)~beliefs can vary with each project, but do not necessarily
correspond with actual evidence in that project''.
Devanbu et al. further  comment that ``programmers give personal experience
as the strongest influence in forming their opinions''. This is a troubling
result, especially given the above comments from Passos and  J{\o}rgensen et al.~\cite{passos11,jorgensen09} about how quickly practitioners form, freeze, and rarely revisit those opinions.

From all we above we conclude that, just as in medicine, 
%our field suffers when
 %software engineers do  not revise old beliefs.  Therefore, 
 it is important for our field
 to regularly  reassess old truisms  like the  delayed issue effect.

\section{Motivation: ``DIE'' is commonly held, yet poorly documented}
\label{sect:belief}
% To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
% understanding of SE theory, then the DIE  may well influence their decisions.
One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
in the SE literature. 
%For example,
%we know that \fig{b81} has been presented to the graduate SE class at North Carolina State University, without quarrel or %critical comment, every year for the last decade.
Yet when we look at the literature, the evidence for
delayed issue effect is both very sparse and very old.

We examined the literature on the delayed issue effect through a combination of snowball sampling~\cite{wohlin2014guidelines} and database search. We began with \cite{boehm2012}, the most recent publication by Barry Boehm (at the time of this paper's first draft) that discusses the delayed issue effect. Boehm cites the seminal \textit{Software Engineering Economics}~\cite{Boehm81}.
%, which is bases the original delayed issue effect on four primary sources in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76}. 
We searched Google Scholar for terms such as ``cost to fix'' and ``defect cost'' and ``software quality cost''. The majority of the search results discuss quality measurements or quality improvement, but relatively few discuss cost-to-fix as a function of when the defect was injected or found. Our literature search is by no means exhaustive, but our search results yielded many popular sources such as~\cite{pressman2005software, boehm01,glass02,endres03}\footnote{These sources have a combined citation count of over 14,500 on Google Scholar}, all of which cite \textit{Software Engineering Economics}~\cite{Boehm81} or its derivative works.

\begin{figure}[!t]
% \begin{center}\includegraphics[width=4in]{img/boehm_cost-to-fix.png}\end{center}
\begin{center}\includegraphics[width=4.5in]{img/boehm81.png}\end{center}
 \caption{Historical cost-to-fix curve. Adapted from~\cite{Boehm81}, p. 40.}\label{fig:cost-to-fix}
 \end{figure}

Figure~\ref{fig:cost-to-fix} shows the DIE as reported in \textit{Software Engineering Economics}~\cite{Boehm81} based on data from large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76}. 
%These studies are most often cited by secondary sources on the delayed issue effect (e.g., \cite{boehm01, mcconnell01,glass02}. 
We note that it is unclear from the text in~\cite{Daly77} and \cite{Boehm76} if cost is defined in terms of effort, or in actual cost (i.e., labor, materiel, travel, etc). The data points from these studies are not published for analysis.
%These seminal studies are the basis for most DIE claims and that the cost to find and fix a defect monotonically increases with lifecycle phase.

In the 40 years since these initial studies, few studies have explored the cost to resolve issues
as a function of lifecycle phase. 
Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects. Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but the raw data points were not provided and thus cannot be confirmed. A goal of agile methods 
is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}. Relatively little empirical data exists on this point.
Elssamadisy and Schalliol~\cite{Elssamadisy02} offer an anecdotal report on the growing, high cost of rework in a 50 person, three-year, 500KLOC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their 
exact figures. This was a common theme in the literature reviewed for this paper-- i.e.  that  it was no longer possible to access the data used to make prior conclusions.

@@

@@

Some studies that report less-than large
 increase in the effort required to fix delayed issues. Royce~\cite{Royce98} studied  a million-line, safety-critical missile defense system. Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly. Boehm~\cite{Boehm10} attributes this success to the  development process, which focused on removing architecture risk early in the development lifecycle. Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects, two student projects of 2000 deliverable source instructions, is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}). Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to fix certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project). Again, the data points from these studies are not available, and thus newer evidence both in favor of and contrary to the DIE cannot be evaluated.

To gain a sense of how current the perception of the DIE is, 
we conducted two surveys of software engineers. 
The surveys collected data on software engineers' views of the DIE and other commonly held software engineering ``laws''. The surveys were conducted using Amazon's Mechanical Turk. The first survey was conducted only with professional software engineers. Participants were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology. The  second survey was conducted with Program Committee members of the ESEC/FSE 2015 and ICSE 2014 conferences solicited via email.

The practitioner survey presented the following law: ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}). We abbreviate this law as RqtsErr.\footnote{We use the RqtsErr formulation since this issue typically needs no supportive explanatory
text. If we had asked respondents about our more general term ``delayed issue
effect'', we would have had to burden our respondents with extra explanations.}
The PC member survey presented the RqtsErr law and an additional law on the DelayedIssueEffect: ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix''.  The respondents answered two questions in response to each law:
\bi
\item \textbf{Agreement:} ``Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree. A text box was provided to explain the answer. 
\item \textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were: -1: I don't know, 0: this law does not apply at all, ..., 5: always applies. Respondents were required to explain the applicability score in a text box.
\ei

\begin{figure}[ht] 
\scriptsize 
 \begin{center}
\begin{tabular}{l|c|c|c|c|c}
 &  & \multicolumn{2}{c|}{agreement} & \multicolumn{2}{c}{applicability} \\
Practitioner survey  & N & med & mode & med & mode \\
\hline 
\textbf{Rqts errors are most expensive...} & 16 & 5 & 5 & 4 & 5 \\ 
Inspections can remove 90\% of defects & 18 & 4 & 5 & 4 & 5 \\
80-20 rule (defects to modules) & 12 & 4 & 5 & 4 & 5 \\
Most time is spent removing errors & 16 & 4 & 4 & 4 & 5 \\ 
Process maturity improves output & 17 & 4 & 4 & 4 & 4 \\ 
Missing reqts are hardest to fix & 17 & 4 & 4 & 4 & 4 \\
Reuse increases prod. and qual. & 16 & 4 & 4 & 4 & 4 \\
OO-programming reduces errors & 13 & 4 & 4 & 4 & 3 \\
Adding manpower to a late project & 15 & 4 & 4 & 4 & 4 \\
Smaller changes have higher error density & 14 & 3 & 3 & 3.5 & 5 \\
A developer is unsuited to test own code & 17 & 3 & 1 & 4 & 5\\
 \multicolumn{6}{l}{}\\
\multicolumn{6}{l}{Researcher survey} \\\hline 
Process maturity improves output & 4 & 4 & 4 & 4 & 5 \\
\textbf{Rqts errors are most expensive...} & 30 & 4 & 4 & 4 & 4   \\ 
\textbf{DelayedIssueEffect} & 30 & 4 & 4 & -- & --  \\
Reuse increases prod. and qual. & 6 & 4 & 4 & 4 & 4 \\
80-20 rule (defects to modules) & 6 & 4 & 4 & 4 & 3 \\
Missing reqts are hardest to fix & 7 & 4 & 4 & 4 & 3 \\
OO-programming reduces errors & 6 & 4 & 4 & 3 & 4 \\
Inspections can remove 90\% of defects & 7 & 4 & 4 & 3 & 3 \\
Adding manpower to a late project & 4 & 3 & 4 & 4 & 3 \\
Most time is spent removing errors & 6 & 3 & 3 & 4 & 4 \\ 
Smaller changes have higher error density & 4 & 3 & -- & 4 & 4 \\
A developer is unsuited to test own code & 7 & 2 & 1 & 3 & 3
\end{tabular} 
 \end{center}
\caption{Agreement and applicability of SE axioms.}
\label{fig:survey_results}
\end{figure}

Both  practitioners and researchers strongly believed in RqtsErr. In both sets of responses, RqtsErr received  scores higher than most
other laws. Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers. From the free response texts, we note that the researchers who disagreed with RqtsErr generally asserted that requirements change can be expensive, but that the effect depends on the process used (e.g., agile vs. waterfall) and the adaptability of the system architecture.
